Elastic Scattering in General Relativistic Ray Tracing for Neutrinos

### 2017-05-26 paper.tex
Add Hebert to acknowledgements
Update paper structure to reflect current project directions
* remove sections related to closure relation (not going to do this now)
* add new test sections

\documentclass[aps,floatfix,prd,superscriptaddress,twocolumn]{revtex4-1}
% Intended to be built with pdflatex, using the REVTEX 4.1 package.
% Template taken from FermiLab APS Jouranl template.

@@

\begin{abstract}
  We present a covariant ray tracing algorithm for computing high-resolution
  neutrino distributions in general numerical spacetimes with hydrodynamical
  sources.
  Our formulation treats elastic scattering of neutrinos (changing direction
  but not energy) off of nuclei and
  nucleons by incorporating estimates of the background neutrino fields.
  These background fields may be taken from a low-order moment simulation,
  be supplied analytically in simple cases,
  or be ignored entirely, in which case the method
  reduces to a standard state-of-the-art ray tracing formulation.
  It handles radiation in regimes spanning optically thick to optically thin.
  We test the new code, highlight its strengths and weaknesses, and
  apply it to simulations of neutron star mergers to
  1) compute neutrino fluxes as a function of energy and emission angle, and
  2) compute the radiation pressure tensor in the funnel of the disk,
  demonstrating a way to improve the closure relations used in the truncated
  moment formalism.
\end{abstract}

@@

\subsection{Outline}
In Sec.~\ref{sec:formulation} we derive the ray tracing equations from the
Boltzmann Equation and describe our numerical scheme.
In Sec.~\ref{sec:tests} we present some tests of the code.
In Sec.~\ref{sec:applications} we present two applications:
neutrino spectra as a function of observer's angle, and
a variable closure relation for a truncated moment formalism.
We present both applications in the dynamical environments following the merger
of a neutron star--neutron star and a black hole--neutron star binary.

@@

\subsection{Dummy Tests}
\label{ssec:tests_basic}
[Maybe we don't need to present these tests, but make sure to establish
confidence in the basics:
gravitational redshift, Doppler shift,
geoesic curvature, thermodynamic equilibrium.]
\todo{setup and run dummy tests}

\subsection{Homogenous Absorbing Star}
\label{ssec:test_ab_star}
[Test described in~\cite[Sec.~3.2]{smit1997-two_moment}.]
\todo{Jerred: run test}

\subsection{Homogenous Scattering Star}
\label{ssec:test_sc_star}
\todo{maybe drop this test}
[Test presented in~\cite{humm1971-grey_transfer}
and described in~\cite[Sec.~9.1.1]{abdi2012-monte_carlo}.
Our version would be a point source embedded in a spherical distribution
of purely elastic-scattering material. It requires an analytical estimate
of $J$.]
\todo{formulate analytic $J$.}

\subsection{Post-Bounce Neutron Core}
\label{ssec:test_bounce_core}
[Standard radiation transport test described in~\cite{ocon2015-gr1d_with_nu},
and in \cite[App.~E.6]{fouc2015-m1_nsbh},
and in~\cite{abdi2012-monte_carlo}.]
\todo{Evan: provide initial data and $H^r(\varepsilon)$}
\todo{Jerred: setup and run test}

\subsection{Disk Comparison to M1}
\label{ssec:test_disk_comparison}
[Compute total and relative number and energy luminosities for a disk evolved
in one of~\cite{fouc2015-m1_nsbh, fouc2016-m1_nsns, fouc2016-m1_evolve_n}.
Expect agreement within factors of a few (?).]
\todo{code up moment integrals}
\todo{compute luminosities from M1}
\todo{compute luminosities from ray tracing}

\begin{figure}
  \includegraphics[width=\columnwidth]{fig-homogeneous_isotropic}
  \caption{
    Equilibrium distribution functions in a homogeneous background at
    $\varrho=6.22 \times 10^{12}\,{\rm g}\,{\rm cm}^{-3}$,
    $T=5$~MeV,
    $Y_e=0.3$,
    using the LS220 equation of state, in which the equilibrium
    neutrino chemical potentials are
    $\mu_{\nu_e}=-\mu_{\bar{\nu}_e}=62.04$~MeV, and 
    $\mu_{\nu_x}=0$.
    The solid lines give the analytic Fermi-Dirac distribution,
    and the points are computed via ray tracing.}
  \label{fig:homogeneous_isotropic}
\end{figure}

@@

\subsection{Emission Spectra}
\label{ssec:spectra}
The merger models are described in \cite{fouc2015-m1_nsbh, fouc2016-m1_nsns}.
[Something like this is presented in~\cite[Fig.~10-11]{pere2014-nu_wind};
compare.]
\todo{code up moment integrals}
\todo{compute $H^r(\theta;\varepsilon)$ in nsns and nsbh}
\todo{maybe compute at several times}

\subsection{Closure Relation}
\label{ssec:closure}
\todo{figure out your approach}
[This is a little tricky. Unless we can come up with a new improved analytic
closure relation and use ray tracing to evaluate its trustworthiness,
it's going to be hard to get anything useful from computing
$S^{\alpha\beta}$.]

[Maybe the variable Eddington closure is more suitable to this application.
If I understand it right, applying ray tracing to that method would require
computing $S^{\alpha\beta}$ on the fly throughout the spatial volume.
That would certainly be doable for a quasi-stationary disk configuration.]
\todo{compute $S^{\alpha\beta}$ and evaluate its effect on evolution}

@@

We use the following procedure:
\begin{enumerate}
\item
  Interpolate the fluid temperature, $T$,
  equilibrium neutrino chemical potential, $\mu_\nu$,
  and evolved neutrino energy and number densities in the fluid frame,
  $J$ and $G$, from the simulation grid.
\item
  Compute the average neutrino energy in the fluid frame
  $\langle \varepsilon \rangle \equiv J/G$.
\item
  Compute the neutrino spectral temperature, assuming equilibrium with the
  fluid
  \begin{equation}
    T_\nu = \langle \varepsilon \rangle
    \frac{\mathscr{F}_2(\mu_\nu/T)}{\mathscr{F}_3(\mu_\nu/T)},
  \end{equation}
  where $\mathscr{F}_b$ is the Fermi integral
  \begin{equation}
    \mathscr{F}_b(\eta) = \int_0^\infty dx \, x^b (1+e^{x-\eta})^{-1}.
  \end{equation}
\item
  Assume the background neutrino fields have the same total density
  as the evolved moments
  \begin{align}
    J(\varepsilon) &=
    J \frac{\varepsilon^3}{T_\nu^4 \,\mathscr{F}_3(\mu_\nu/T)}
    (1+\exp(\varepsilon/T_\nu-\mu_\nu/T)^{-1}, \\
    H^\mu(\varepsilon) &=
    H^\mu \frac{\varepsilon^3}{T_\nu^4\, \mathscr{F}_3(\mu_\nu/T)}
    (1+\exp(\varepsilon/T_\nu-\mu_\nu/T)^{-1}.
  \end{align}
\end{enumerate}

@@

\subsubsection*{Acknowledgements}
This research is supported in part by the Joint Institute for Nuclear
Astrophysics---Center for the Evolution of the Elements (JINA-CEE)
and the National Science Fondation under grant numbers X. \todo{grant nos?}
We thank Sherwood Richers for providing his neutrino transport code
\lstinline{Sedonu}~\footnote{\url{https://bitbucket.org/srichers/sedonu}}
for comparisons.
We thank Luke Roberts and Yonglin Zhu for helpful discussions.

>>>>

\documentclass[aps,floatfix,prd,superscriptaddress,twocolumn]{revtex4-1}
% Intended to be built with pdflatex, using the REVTEX 4.1 package.
% Template taken from FermiLab APS Journal template.

@@

\begin{abstract}
  We present a covariant ray tracing algorithm for computing high-resolution
  neutrino distributions in general numerical spacetimes with hydrodynamical
  sources.
  Our formulation treats elastic scattering of neutrinos (changing direction
  but not energy) off of nuclei and
  nucleons by incorporating estimates of the background neutrino fields.
  These background fields may be taken from a low-order moment simulation
  or be ignored entirely, in which case the method
  reduces to a standard state-of-the-art ray tracing formulation.
  It handles radiation in regimes spanning optically thick to optically thin.
  We test the new code, highlight its strengths and weaknesses, and
  apply it to simulations of neutron star mergers to
  compute neutrino fluxes as a function of energy and emission angle.
\end{abstract}

@@

\subsection{Outline}
In Sec.~\ref{sec:formulation} we derive the ray tracing equations from the
Boltzmann Equation and describe our numerical scheme.
In Sec.~\ref{sec:tests} we present some tests of the code.
In Sec.~\ref{sec:applications} we present an application:
computing neutrino spectra as a function of observer's angle
in the dynamical environment following the merger of two neutron stars.

@@

\subsection{Homogeneous Equilibrium}
\label{ssec:test_equilibrium}
Homogeneous matter, optically thick. Plot in Fig.~\ref{fig:homogeneous_isotropic}.
\todo{describe test}

\subsection{Doppler Shift}
\label{ssec:test_doppler}
Homogeneous matter moving with v in +z direction. Stationary observer.
\todo{add plot of $\varepsilon(\mu)$}

\subsection{Gravitational Redshift and Geodesic Curvature}
\label{ssec:test_gravity}
Optically thick star in Schwarszchild.
todo{describe $L(r)$, as a function of lapse}
optically thick disk in Kerr (Fishbone-Moncrief disk):
set up to give two different surfaces for nue vs nua, but same temp (?).
\todo{plot of analytic and numerical geodesics}
\todo{plot of $G$ nue vs nua}

\subsection{Homogenous Absorbing Star}
\label{ssec:test_ab_star}
[Test described in~\cite[Sec.~3.2]{smit1997-two_moment}.]
\todo{run test for optically thick and thin}

\subsection{Massive Collapsed Star with Scattering Envelope}
\label{ssec:test_scat_envelope}
[Standard radiation transport test described in~\cite{ocon2015-gr1d_with_nu},
and in \cite[App.~E.6]{fouc2015-m1_nsbh},
and in~\cite{abdi2012-monte_carlo}.]
\todo{add plots of $\langle \varepsilon \rangle(\mu)$}

\subsection{Disk Comparison to M1}
\label{ssec:test_disk_comparison}
[Compute total and relative number and energy luminosities for a disk evolved
in one of \cite{fouc2015-m1_nsbh, fouc2016-m1_nsns, fouc2016-m1_evolve_n}.
Expect agreement within factors of a few (?).]
\todo{compute luminosities from M1 and RT, present table}

\begin{figure}
  \includegraphics[width=\columnwidth]{fig-homogeneous_isotropic}
  \caption{Equilibrium distribution functions in a homogeneous background at
    $\varrho=6.22 \times 10^{12}\,{\rm g}\,{\rm cm}^{-3}$,
    $T=5$~MeV,
    $Y_e=0.3$,
    using the LS220 equation of state, in which the equilibrium
    neutrino chemical potentials are
    $\mu_{\nu_e}=-\mu_{\bar{\nu}_e}=62.04$~MeV, and 
    $\mu_{\nu_x}=0$.
    The solid lines give the analytic Fermi-Dirac distribution,
    and the points are computed via ray tracing.
    Described in Sec.~\ref{ssec:test_equilibrium}}
  \label{fig:homogeneous_isotropic}
\end{figure}

@@

\subsection{Emission Spectra}
\label{ssec:spectra}
The merger models are described in \cite{fouc2015-m1_nsbh, fouc2016-m1_nsns}.
Something like this is presented in \cite[Figs.~10-11]{pere2014-nu_wind};
compare.]
\todo{compute $H^r(\mu;\varepsilon)$ in nsns and nsbh}
\todo{compute $\langle \varepsilon \rangle(\mu)$}

@@

We use the following procedure:
\begin{enumerate}
\item
  Interpolate the fluid temperature, $T$,
  equilibrium neutrino chemical potential, $\mu_\nu$,
  and evolved neutrino energy and number densities in the fluid frame,
  $J$ and $G$, from the simulation grid.
\item
  Compute the average neutrino energy in the fluid frame
  \begin{equation}
    \langle \varepsilon \rangle \equiv J/G.
  \end{equation}
\item
  Compute the neutrino spectral temperature, assuming equilibrium with the
  fluid
  \begin{equation}
    T_\nu = \langle \varepsilon \rangle
    \frac{\mathscr{F}_2(\mu_\nu/T)}{\mathscr{F}_3(\mu_\nu/T)},
  \end{equation}
  where $\mathscr{F}_b$ is the Fermi integral
  \begin{equation}
    \mathscr{F}_b(\eta) = \int_0^\infty dx \, x^b (1+e^{x-\eta})^{-1}.
  \end{equation}
\item
  Assume the background neutrino fields have the same total density
  as the evolved moments
  \begin{align}
    J(\varepsilon) &=
    J \frac{\varepsilon^3}{T_\nu^4 \,\mathscr{F}_3(\mu_\nu/T)}
    (1+\exp(\varepsilon/T_\nu-\mu_\nu/T)^{-1}, \\
    H^\mu(\varepsilon) &=
    H^\mu \frac{\varepsilon^3}{T_\nu^4\, \mathscr{F}_3(\mu_\nu/T)}
    (1+\exp(\varepsilon/T_\nu-\mu_\nu/T)^{-1}.
  \end{align}
\end{enumerate}

@@

\subsubsection*{Acknowledgements}
This research is supported in part by the Joint Institute for Nuclear
Astrophysics---Center for the Evolution of the Elements (JINA-CEE)
and the National Science Fondation under grant numbers X. \todo{grant nos?}
We thank Sherwood Richers for providing his neutrino transport code
\lstinline{Sedonu}~\footnote{\url{https://bitbucket.org/srichers/sedonu}}
for comparisons.
We thank Luke Roberts and Yonglin Zhu for helpful discussions.
We are indebted to Fran\c{c}ois H\'{e}bert for his timely help in finding and
correcting bugs in the time-interpolation algorithm.
